---
title: "Storm Data Analysis"
author: "Dave Patterson"
date: "September 13, 2015"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis
The purpose of this report is to explore the NOAA Storm Database and analyze which types of storm events 
1) are most harmful with respect to population health and 
2) which types of events have the greatest economic consequences?
This analysis should help to prioritize resources to prepare for severe weather events.

## Data Processing

The following code blocks loads the required packages, downloads a copy of the 
U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from the 
course repository for all the storms in the U.S. from 1950 through 2011.

```{r load libs}
packages <- c("data.table", "dplyr", "lubridate", "stringr", "ggplot2")
sapply(packages, require, character.only = TRUE, quietly = TRUE)
```

```{r get files}
original_data  <- "http://www.ncdc.noaa.gov/stormevents/details.jsp"
storm_data_url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
storm_data_doc <- "http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
storm_data_faq <- "http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
storm_data_bz2 <- "StormData.csv.bz2"
storm_data_events_file <- "StormDataEventTable.csv"

# if the data file is not already present, get the source data file
if(!(file.exists(storm_data_bz2))) {download.file(storm_data_url, destfile = storm_data_bz2)}
```

Next, we read in the data and prepare it for analysis. Given the large volume of data 
(over 900,000 storm events), we only read in the data needed to answer the two questions 
of interest to speed up the processing of the data. In the list below, "NULL" indicates 
that data field is ignored in this analysis.

```{r get headers}
# grab the headers
storm_data_headers <- names(read.csv(storm_data_bz2, header = T, nrow = 1))
# create a vector to ignore most of the variables when we read them in since we're not
# interested in most of them -- only in data the can affect the answers
# i.e. storm data that impacts health or property
header_char_classes <- rep("NULL", 37)
# set the variables we are interested in to type char so they read in quickly
header_char_classes[c(2,8,23:28,36:37)] <- "character"
# verify the mapping
cbind(storm_data_headers, header_char_classes)
```

```{r read data}
```

```{r analyze event types}
# explore the data
```

```{r clean events}
# look for outliers that may distort the results; correct or exclude
```

```{r health impact}
```

```{r financial impact}
```

