---
title: "Financial and Health Impacts of U.S. Storms From 1996 Through 2011"
author: "Dave Patterson"
date: "September 19, 2015"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis
The purpose of this report is to explore the NOAA Storm Database and analyze the types of 
storm events to understand  
1. which are most harmful with respect to population health and  
2. which have the greatest economic consequences?  
This analysis should help to prioritize resources to prepare for severe weather events.

Prior to 1996, there was no standard classification of storm types. Effective beginning in 1996, 
storm events were grouped into 48 standard event types. Prior to that, the storm data was 
both sparse and inconsistent. In addition, due to inflation, the value of money prior to 1996
in comparison to today was significantly great enough to potentially skew the data. 
For these reasons, we will only work with the last 15 years of the data available. 

## Data Processing

A copy of the United States National Oceanic and Atmospheric Administration's (NOAA) 
storm database was downloaded from the John Hopkins Data Science Coursera course 
repository for all the storms in the United States from 1950 through 2011. 
The data was downloaded September 19, 2015.

```{r load libs}
packages <- c("data.table", "dplyr", "lubridate", "stringr", "ggplot2", "cowplot")
sapply(packages, require, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)
```

```{r get files}
original_data  <- "http://www.ncdc.noaa.gov/stormevents/details.jsp"
storm_data_url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
storm_data_doc <- "http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
storm_data_faq <- "http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
storm_data_bz2 <- "StormData.csv.bz2"
storm_data_events_file <- "StormDataEventTable.csv"

# if the data file is not already present, get the source data file
if(!(file.exists(storm_data_bz2))) {download.file(storm_data_url, destfile = storm_data_bz2)}
```

Given the large volume of data 
(over 900,000 storm events), we only read in the data needed to answer the two questions 
of interest. This will eliminate two thirds of the data fields and simplify, 
as well as speed up, the processing of the data. In the list below, "NULL" indicates 
that data field is ignored in this analysis.

```{r select variables}
# grab the headers
storm_data_headers <- names(read.csv(storm_data_bz2, header = T, nrow = 1))
# create a vector to ignore most of the variables when we read them in since we're not
# interested in most of them -- only in data the can affect the answers
# i.e. storm data that impacts health or property
header_char_classes <- rep("NULL", 37)
# set the variables we are interested in to type char so they read in quickly
header_char_classes[c(2,8,23:28,36:37)] <- "character"
# verify the mapping
cbind(storm_data_headers, header_char_classes)
```

The data is loaded and prepared for analysis. Three levels of filters applied:
1. Date: only the storm data collected after January 1, 1996 is used
2. Fields: only the fields needed to answer the questions, as shown in the table above, is used
3. Impact: only storm events that had a health or financial impact is used

The date filter still leaves nearly two thirds of a million storm events, many of which 
have no impact on the questions we are looking to answer. The impact filter eliminates 
more than two thirds of the storm events in the 15 year period, but still leaves a bit 
over 200,000 storm events, all of which have a bearing on the two questions.

```{r read and filter data, cache=TRUE}
# read in only the variables we're interested in and only the data after the new
# stardard has gone into effect and that has an impact on the questions
# Ignoring the early data removes the most noise and about a third of the data 
storm_data <- tbl_df(read.csv(storm_data_bz2, 
                              stringsAsFactors=FALSE, 
                              # filter the fields
                              colClasses = header_char_classes))  %>% 
              # filter by date
              mutate(BGN_DATE =  mdy_hms(BGN_DATE)) %>% 
              filter(BGN_DATE >= "1996-01-01") %>%
              # coerce the numeric data to be the correct type
              mutate(FATALITIES = as.integer(FATALITIES),
                     INJURIES = as.integer(INJURIES),
                     PROPDMG = as.numeric(PROPDMG),
                     CROPDMG = as.numeric(CROPDMG),
                     REFNUM = as.integer(REFNUM)) %>%
              # finally keep only the storm data that can affect the questions; this removes
              # about two thirds of the remaining data, none of which helps answer the questions
              # since we're looking at totals, not averages
              filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)
```

In a first look at the storm data, we can see the storm event types do not always align 
with the standard event names.

```{r event names}
# read in the standard event types used starting 1996
# extracted from storm_data_doc file (Table 1. Storm Data Event Table)
standard_events <- read.csv(storm_data_events_file, comment.char = "#", stringsAsFactors = FALSE)

# and from our storm data
storm_data_events <- unique(storm_data$EVTYPE)
```

The storm data contains **`r length(storm_data_events)`** unique storm event types in contrast 
to the `r length(standard_events$Event.Name)` standard event types. It would greatly simplify the
analysis if the `r length(storm_data_events)` unique storm event types could be grouped into the
`r length(standard_events$Event.Name)` standard event types. Since the filtered data contains 
`r length(storm_data$EVTYPE)` storm events, it is obvious that we'll need some tools to help with 
the analysis. These functions were useful in exploring and cleaning the storm data.

```{r useful functions}
# define a function to select storm events from the data set based on regular expressions
# this aids in generating regexps to group events into std types
get_event <- function(src, pattern) {
               unique(grep(pattern, 
                           src, 
                           ignore.case = TRUE, 
                           value = TRUE))}
# and an example call
# get_event(events, "^\\s*(tstm|thunder)")

# define a function to get the remarks for a specific storm reference
get_remarks_for <- function(ref) {
  storm_data[(storm_data$REFNUM == ref),"REMARKS"]
}

# define a function to replace the dirty event types with standard names
# NOTE that argument order is set to allow pipelining
replace_event <- function(src, pattern, replacement) {
                   idx <- grepl(pattern, src$EVTYPE, ignore.case = TRUE)
                   src[idx, "EVTYPE"] <- replacement
                   src}

# define a function to replace the dirty event types with standard names while
# excluding a specific phrase or word
# NOTE that argument order is set to allow pipelining
replace_event_wo <- function(src, pattern, exclude, replacement) {
                   idx <- grepl(pattern, src$EVTYPE, ignore.case = TRUE) &
                          !grepl(exclude, src$EVTYPE, ignore.case = TRUE)
                   src[idx, "EVTYPE"] <- replacement
                   src}

```

Using these functions the storm data events can be mapped into the standard events.

```{r map events, cache=TRUE}
# Use regular expressions to map the dirty event type fields into standard names
storm_data <- select(storm_data, c(1:8,10)) %>% 
  replace_event("^Astronomical Low Tide", 
                "Astronomical Low Tide") %>%
  replace_event("(coast|cstl|beach).*(flood|erosion)", 
                "Coastal Flood") %>%
  replace_event("^(extended )*cold", 
                "Cold/Wind Chill") %>%
  replace_event("^(land|mud|rock).*(slide|slump)", 
                "Debris Flow") %>%
  replace_event_wo("fog", 
                   "freezing", 
                   "Dense Fog") %>%
  replace_event_wo("dust", 
                   "devil", 
                   "Dust Storm") %>%
  replace_event("dust devil", 
                "Dust Devil") %>%
  replace_event("^(heat wave|(excessive|record) heat)", 
                "Excessive Heat") %>%
  replace_event("^((extreme|unseason).*(cold|chill)|hyp(o|er)thermia)", 
                "Extreme Cold/Wind Chill") %>%
  replace_event("^\\s*flash.*flood", 
                "Flash Flood") %>%
  replace_event("^(((ice jam|river)*\\s*flood)|urban.*fld|high water)", 
                "Flood") %>%
  replace_event_wo("(freez|frost)", 
                   "fog", 
                   "Frost/Freeze") %>%
  replace_event("^(funnel|whirlwind)", 
                "Funnel Cloud") %>%
  replace_event("^(small )?hail", 
                "Hail") %>%
  replace_event("(^heat$|warm)", 
                "Heat") %>%
  replace_event("^((heavy|hvy|torrential|unseasonal).*rain|rain)", 
                "Heavy Rain") %>%
  replace_event("^((heavy|excessive).*snow|snow squall)", 
                "Heavy Snow") %>%
  replace_event("^\\s*(high|heavy|rough|hazardous).*(surf|seas|swell)", 
                "High Surf") %>%
  replace_event("^Hurricane|Typhoon", 
                "Hurricane (Typhoon)") %>%
  replace_event_wo("(ic[ey]|glaze)", 
                   "jam", 
                   "Ice Storm") %>%
  replace_event("^lake.*snow", 
                "Lake-Effect Snow") %>%
  replace_event("^Marine (Thunderstorm|Tstm) Wind", 
                "Marine Thunderstorm Wind") %>%
  replace_event("^rip current", 
                "Rip Current") %>%
  replace_event("^Seiche|(astro.*)?high tide", 
                "Seiche") %>%
  replace_event("^(storm surge|coast.*storm|tidal|rogue|high tide)", 
                "Storm Surge/Tide") %>%
  replace_event_wo("(Strong Wind|wind damage|(down|micro)burst)", 
                   "marine",
                   "Strong Wind") %>%
  replace_event("^\\s*(tstm|thunder).*", 
                "Thunderstorm Wind") %>%
  replace_event_wo("wind", 
                   "chill|marine|strong|thunder", 
                   "High Wind") %>%
  replace_event("^(water|land)spout", 
                "Waterspout") %>%
  replace_event("fire", 
                "Wildfire") %>%
  replace_event_wo("((wint[ery]+ (weather|mix))|snow|mixed precip)", 
                   "heavy|excessive|lake", 
                   "Winter Weather") %>%
  replace_event("other|marine accident|drowning", "Other") %>%
  mutate(EVTYPE = as.factor(str_to_title(EVTYPE)))

other_events <- storm_data$EVTYPE[storm_data$EVTYPE == "Other"]
```

## Results

Having mapped the storm event data into standard type names, we can see the frequency of 
each of the types of events. Note that there are `r length(unique(storm_data$EVTYPE))` 
event types in our data. The extra "Other" category is the catchall for those 
`r length(other_events)` events that were reported, but did not seem to 
fall into one of the standard types. 
"Other" includes these event types: "Other", "Marine Accident", and "Drowning". 
Thunderstorm Wind is the most common event by an order of magnitude.

```{r type table, cache=TRUE}

# Show the table of counts of all event types
table(storm_data$EVTYPE)

# function to plot horizontal barchart of top ten events most impactful events for an impact
# evdata should be a "tbl_df" of ten rows, with the name of the storm event as the first column
# and the count of the impact (frequency, fatalities, injuries, or damage) as the second column
tt_plot <- function(evdata, title, cnt_label) {
  evdata %>% 
    setnames(c("Event", "Count")) %>%
    mutate(Event = factor(as.character(Event), levels=Event)) %>%
    ggplot(aes(x = desc(Event),
               y = Count, 
               fill = Event)) +
           geom_bar(stat="identity") + 
           scale_x_continuous(breaks=(-1:-10), labels=evdata$Event) +
           guides(fill=FALSE) + 
           ggtitle(title) +
           ylab(cnt_label) +
           coord_flip() +
           theme_bw() +
           theme(axis.title.y = element_blank())
}

# Plot the ten most frequent events (tte = top ten events)
tte <- head(sort(table(storm_data$EVTYPE), decreasing = TRUE), 10) %>% 
       as.data.table(keep.rownames = TRUE)
tt_plot(tte, "Top Ten Most Frequent Events", "Occurrences")

```

### Health Impacts

Here are the top ten storm event types that have an impact on public health, broken out by 
Fatalities and Injuries.

```{r health impact, cache=TRUE}

# top ten events causing the most fatalities
ttf <- tbl_df(aggregate(FATALITIES ~ EVTYPE, storm_data, sum)) %>% 
    arrange(desc(FATALITIES)) %>% 
    head(10)
ttfp <- tt_plot(ttf, "Top Ten Most Fatal Events", "Fatalities")

# top ten events causing the most injuries
tti <- tbl_df(aggregate(INJURIES ~ EVTYPE, storm_data, sum)) %>% 
    arrange(desc(INJURIES)) %>% 
    head(10)
ttip <- tt_plot(tti, "Top Ten Most Injurious Events", "Injuries")

# Print both graphs as a single plot
plot_grid(ttfp, ttip, nrow = 2)
```

### Financial Impacts

Here are the top ten storm event types that have a financial impact, broken out by 
property damage and crop damage.

```{r financial impact, cache=TRUE}

# Apply the exponents to the damage costs, then total by event types
# What is the multiplier of the EXP variables
## Convert to billions of dollars
get_exp_val <- function (exp_var) {
  switch(exp_var, "K" = 10^-6,
                  "M" = 10^-3,
                  "B" = 1,
                  10^-9)}

# normalize the asset damage
storm_data <- storm_data %>%
       mutate(PROPDMGNORM = PROPDMG * sapply(PROPDMGEXP, get_exp_val)) %>%
       mutate(CROPDMGNORM = CROPDMG * sapply(CROPDMGEXP, get_exp_val)) %>%
       group_by(EVTYPE)

# top ten events causing the most property damage
ttp <- tbl_df(aggregate(PROPDMGNORM ~ EVTYPE, storm_data, sum)) %>% 
    arrange(desc(PROPDMGNORM)) %>% 
    head(10)
ttpp <- tt_plot(ttp, "Top Ten Events Causing the Most Property Damage", "Billions of Dollars Damage")

# top ten events causing the most crop damage
ttc <- tbl_df(aggregate(CROPDMGNORM ~ EVTYPE, storm_data, sum)) %>% 
    arrange(desc(CROPDMGNORM)) %>% 
    head(10)
ttcp <- tt_plot(ttc, "Top Ten Events Causing the Most Crop Damage", "Billions of Dollars Damage")

# Print both graphs as a single plot
plot_grid(ttpp, ttcp, nrow = 2)
```

