---
title: "ToothGrowth Analysis"
author: "Dave Patterson"
date: "November 21, 2015"
output: pdf_document
---

## Overview

This data shows the effect of Vitamin C on tooth growth in guinea pigs.

## Data Summary

```{r load libs}
packages <- c("datasets", "data.table", "dplyr", "ggplot2")
sapply(packages, require, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)
```

The data set  reflects the length of teeth in each of 10 guinea pigs at each of
three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery
methods (orange juice or ascorbic acid).

```{r basic summary}
tg.df <- tbl_df(ToothGrowth)

str(tg.df)
summary(tg.df)
table(tg.df$supp, tg.df$dose)
```

Plotting the data reveals that the two delivery mthods have different impacts at
different dosages. In partlicular, asorbic acid appears to be less effective
than orange juice except at the highest dose where it appears to be more
effective for increasing tooth length. See the Appendix for a boxplot suggesting
this.

## Hypothesis

## Conclusions

\pagebreak

## Appendix

```{r boxplot}
ggplot(tg.df, aes(x = dose, y = len)) + 
  ggtitle("Tooth Length by Dose and Delivery") +
  labs(x = "Dose", y = "Tooth Length") +
  geom_boxplot(aes(x = as.factor(dose), fill = supp), position = "dodge") +
  scale_fill_discrete(name="Delivery\nMechanism",
                      labels=c("Orange\nJuice", "Ascorbic\nAcid"))
```
