---
title: "Exploring the Central Limit Theorem with the Exponential Distribution"
author: "Dave Patterson"
date: "November 18, 2015"
output: 
  pdf_document: 
    toc: yes
  html_document: default
---

## Overview: 

I will investigate the exponential distribution and compare it with the Central 
Limit Theorem. To do this, I will compare some sample means and variances to
their theoretical values. I will also compare the distribution of a large
collection of random exponentials and the distribution of a large collection of
averages of 40 exponentials.

## Simulations: 
One simulation data run was used for consistency across the plots. The data
consist of 1000 iterations of a set of 40 samples of the random exponential
distribution, using `rexp`, with a lambda of 1/5. Plots were created for
exploring the means, the variance, and the central limit theorem. The
`sim_param` function runs the requested function (`mean`, `var`, or `clt_func`)
on the data set.

```{r simulations, cache=TRUE}

# generate the simulation data of random exponentials samples
lambda <- 0.2
iterations <- 1000
sample.size <- 40
rexp.mat  <- matrix(rexp(iterations * sample.size, lambda), iterations)

# sim_param returns a dataframe that has a function applied to the sample data 
sim_param <- function (func, dat = rexp.mat, ...) {
  data.frame(x = c(apply(dat, 1, func, ...)))
}

means <- sim_param(mean)$x
vars  <- sim_param(var)$x
```

## Sample Mean versus Theoretical Mean: 
```{r means comparison}
exp.mean   <- 1 / lambda
means.ave  <- mean(means)
```

The theoretical mean of `rexp` is 1 / lambda, in this case **`r exp.mean`**. The
average mean of our simulation is **`r round(means.ave, 5)`**, which is very
close to the theoretical mean. In the Simulation Means histogram in the
appendix, the simulation average mean is virtually eclipsed by the theoretical
mean.

## Sample Variance versus Theoretical Variance: 

```{r variances comparison}
exp.var   <- (1 / lambda)^2
vars.ave  <- mean(vars)
```

The theoretical variances of `rexp` is the standard deviation squared, in this
case **`r exp.var`**. The average variance of our simulation is 
**`r round(vars.ave, 5)`**, which is again very close to the theoretical
variance, as can be seen in the Appendix.

## Distribution: 

If we apply the central limit theorem to the distribution of means from
the simulation and overlay it with a standard normal curve, we can see that the
distribution of the simulation means are approximately normal.

\pagebreak

## Appendix:
### Code and Plots for Simulation Means
```{r means plot}
my_hist <- function(param, label, theoretical, simulation) {
  hist(param, breaks = 50,
       main = paste("Histogram of Simulation", label),
       col = "lightblue", 
       xlab = label)
  abline(v = simulation,  lw = 3, col = "red")
  abline(v = theoretical, lw = 3, col = "seagreen")
  legend("topright", c("Simulation Average", "Theoretical"), 
         lwd=3,  col=c("red", "seagreen"))
}

my_hist(means, "Means", exp.mean, means.ave)
```

\pagebreak

### Code and Plots for Simulation Variance

```{r variances plot}
my_hist(vars, "Variances", exp.var, vars.ave)
```

\pagebreak

### Code and Plots for Comparing Simulation Means to CLT

```{r distributions comparison}
require(ggplot2)

# define the central limit theorem function
exp.sd <- 1 / lambda
clt_func <- function(x, n) sqrt(n) * (mean(x) - exp.mean) / exp.sd
# and apply it to the same simulation data
clt.dat  <- sim_param(clt_func, dat = rexp.mat, sample.size)

clt.mean <- mean(clt.dat$x)
ggplot(clt.dat, aes(x = x)) +
  ggtitle("Histogram of Simulation Means and CLT") + 
  xlab("Means") + ylab("Density") +
  geom_histogram(binwidth = 0.2, aes(y = ..density..), fill = "lightblue", color = "black") +
  geom_vline(aes(xintercept=clt.mean), color = "red") +
  stat_function(fun = dnorm, size = 2)
```
